A Documentary History of \\ the Nock Combinator Calculus

### 2025-08-28 mss.tex
Post

\newcommand{\authorname}{N. E. Davis~\patp{lagrev-nocfep},\\Curtis Yarvin~\patp{sorreg-namtyv}}
\newcommand{\affiliation}{Zorp Corp, Urbit Foundation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
            %   Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} II:1 (2025):  155–189. \\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
% \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
% \fancyhead[RO]{A History of Nock}
% \fancyfoot[LE,RO]{\thepage}

@@

Perhaps better put, Nock is a family of combinator calculi that sequentially converge on an ``optimal'' expressiveness for certain design desiderata.  This includes an economy of expression (thus several macro opcodes) and consideration of how a higher-level language would invoke stored procedural expressions.  Furthermore an opcode exists which produces and then ignores a computation, intended to signal to a runtime layer that a side effect may be desired by the caller.\footnote{An able if dated document from January 2010, \href{https://github.com/cgyarvin/urbit/blob/1bc9086a3fd62594e8bd631744cb9bf804f6a43b/Spec/urbit/5-whynock.txt}{\texttt{5-whynock.txt}}, further expounds desiderata for Nock in the context of Urbit as operating function.}

@@

The \texttt{U} specification is in some ways the single most interesting historical document of our series.  Yarvin particularly identified a desire to avoid baking abstractions like variables and functions into the \texttt{U} cake, and an emphasis on client–server semantics.  The scry namespace appears \emph{avant la lettre} as a referentially transparent immutable distributed namespace.  \texttt{U} expresses a very ambitious hyper-Turing operator, acknowledging that its own instantiation from the specification is impossible and approximate.  Yarvin grapples in \texttt{U} with the halting problem (via his follow operator) and with the tension between a specification and an implementation (a gulf he highlighted as a human problem in his 2025 LambdaConf keynote address).  Furthermore, asides on issues like the memory arena prefigure implementation details of Vere as a runtime.

@@

\noindent
More concretely, Watt (the predecessor to Hoon) is defined as:

@@

While deviations from the trunk line of the Nock family have been proposed at various points,\footnote{Notably, Ax (see \patp{barpub-tarber}'s Ax, pp.~191–227 in this volume), \textsc{skew}, and \textsc{plan}.} Nock itself has remained the definitional substrate of Urbit since its inception.  It has also been adopted as the primary \textsc{isa} of Nockchain and the NockApp ecosystem.

Why, then, do we contemplate further changes?  The \textsc{skew} proposal by \patp{siprel} and \patp{little-ponnys} argued that Nock 4K represented an undesirable saddle point in the design space of possible Nocks, itself a ``ball of mud'' \citep{Skew2020}.  While \textsc{skew} itself was not adopted, it inspired the development of Plunder and \textsc{plan} as a solid-state computing architecture sharing some ambitions with Urbit and Nock \citep{Plunder2025}.  A rigorously æsthetic argument can thus be sustained that Nock is not yet ``close enough'' to its final, diamond-perfect form to be a viable candidate.

@@

A21: Prefer mechanical simplicity to mathematical simplicity.  Often mechanical simplicity and mathematical simplicity go together.

@@

The most recent version, Nock 4K, appears to provide all of the opcodes necessary for correct and efficient evaluation.\footnote{Modulo the vagaries of the von Neumann architecture, etc.}  It is likely that future versions of Nock will be based genetically on Nock 4K, but with some changes to improve its performance and usability.  The road to zero kelvin is likely very long still, given an abundance of caution, but it also appears to be straight.

\selectlanguage{USenglish}
\printbibliography
\end{document}

>>>>

\newcommand{\authorname}{N. E. Davis~\patp{lagrev-nocfep},\\Curtis Yarvin~\patp{sorreg-namtyv}}
\newcommand{\affiliation}{Zorp Corp, Urban Tiger}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
            %   Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} II:1 (2025):  155–190. \\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
% \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
% \fancyhead[RO]{A History of Nock}
% \fancyfoot[LE,RO]{\thepage}

@@

Put perhaps better, Nock is a family of combinator calculi that sequentially converge on an ``optimal'' expressiveness for certain design desiderata.  This includes an economy of expression (thus several macro opcodes) and consideration of how a higher-level language would invoke stored procedural expressions.  Furthermore an opcode exists which produces and then ignores a computation, intended to signal to a runtime layer that a side effect may be desired by the caller.\footnote{An able if dated document from January 2010, \href{https://github.com/cgyarvin/urbit/blob/1bc9086a3fd62594e8bd631744cb9bf804f6a43b/Spec/urbit/5-whynock.txt}{\texttt{5-whynock.txt}}, further expounds desiderata for Nock in the context of Urbit as operating function.}

@@

The \texttt{U} specification is in some ways the single most interesting historical document of our series.  Yarvin particularly identified a desire to avoid baking abstractions like variables and functions into the \texttt{U} cake, and an emphasis on client–server semantics.  The scry namespace appears for the first time as a referentially transparent immutable distributed namespace.  \texttt{U} expresses a very ambitious hyper-Turing operator, acknowledging that its own instantiation from the specification is impossible and approximate.  Yarvin grapples in \texttt{U} with the halting problem (via his follow operator) and with the tension between a specification and an implementation (a gulf he highlighted as a human problem in his 2025 LambdaConf keynote address).  Furthermore, asides on issues like the memory arena prefigure implementation details of Vere as a runtime.

@@

\noindent
More concretely, Watt (the predecessor to Hoon, \citet{Yarvin2010a}) is defined as:

@@

While deviations from the trunk line of the Nock family have been proposed at various points,\footnote{Notably, Ax (see \patp{barpub-tarber}'s Ax, pp.~191–227 in this volume), \textsc{skew}, and \textsc{plan}.} Nock itself has remained the definitional substrate of Urbit since its inception.  It has also been adopted as the primary \textsc{isa} of the NockApp ecosystem.

Why, then, do we contemplate further changes?  The \textsc{skew} proposal spearheaded by \patp{little-ponnys} and \patp{siprel} (2020) argued that Nock 4K represented an undesirable saddle point in the design space of possible Nocks, itself a ``ball of mud''.  While \textsc{skew} itself was not adopted, it inspired the development of Plunder and \textsc{plan} as a solid-state computing architecture sharing some ambitions with Urbit and Nock \citep{Plunder2025}.  A rigorously æsthetic argument can thus be sustained that Nock is not yet ``close enough'' to its final, diamond-perfect form to be a viable candidate.

@@

@@

The most recent version, Nock 4K, appears to provide all of the opcodes necessary for correct and efficient evaluation.\footnote{Modulo the vagaries of the von Neumann architecture, etc.}  It is likely that future versions of Nock will be based genetically on Nock 4K, but with some changes to improve its performance and usability.  The road to 0K is likely very long still, given an abundance of caution, but it also appears to be straight.

\nocite{Nock13K, Nock12K, Nock11K, Nock10K, Nock9K, Nock8K, Nock7K, Nock6K, Nock5K, Nock4K}
\selectlanguage{USenglish}
\printbibliography
\end{document}
